---
title: "Exploratory Data Analysis - DST Harmonization Wheat Data"
author: "Alliance of Bioversity Internation & CIAT"
date: "2024-02-14"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### 1. Load required packages

```{r, message=FALSE, warning=F}
library(tidyverse)
library(sf)
library(terra)
library(klippy)
library(geodata)
library(plyr)
library(tmap)
library(tmaptools)
library(lme4)
library(simr)
```


#### 2.Structure of the data 

```{r}
wheat <- read.csv("wheat_dst_team_shared_v2.csv", header = T, sep = ",", fileEncoding="latin1")
```

See the dimensions of the data
```{r}
dim(wheat)
```

View columns of the data
```{r}
colnames(wheat)
```

See the structure of the data
```{r}
str(wheat)
```

See some observations of the whole dataset
```{r}
hed_wheat <- head(wheat,3)
rmarkdown::paged_table(hed_wheat, options=NULL) 
```

Let's check the number of NA values in each column
```{r}
wheat_sumr <- wheat |>
  summarise_all(~sum(is.na(.)))
rmarkdown::paged_table(wheat_sumr, options=NULL) 
```

From the NA values summary table we have only (16701 - 12621) values of planting and harvesting date observations

#### 3. Data joining and creating combined columns

```{r}
eth_dist <- geodata::gadm(path = ".", "eth", level = 3) |> st_as_sf()
wheat_pts <- terra::vect(wheat, geom = c('long', 'lat'), crs = "epsg:4326") |> st_as_sf()
dim(wheat_pts)
wheat_joined <- st_join(wheat_pts, eth_dist) 
wheat_joined <- st_coordinates(wheat_joined) |> cbind(as.data.frame(wheat_joined))  |> dplyr::select(unique_id, X, Y, year, source, treatment, n_kgpha, p_kgpha, k_kgpha, grain_yield_kgpha, pl_date_ec, hv_date_ec, NAME_1, NAME_2, NAME_3)
head(wheat_joined, 3)
```

Let's create a trial id by concatenating long, lat, year and source. Besides we round the treatment values and concatenate to create a unique treatment id.

```{r}
wheat_joined <- wheat_joined |> dplyr::mutate(NAME_1 = replace(NAME_1, NAME_1 == "Southern Nations, Nationalities", "SNNP"))|> dplyr::mutate(trial_id = paste(X, Y, year, source, sep = ",")) |> dplyr::mutate(n_rate2 = round_any(n_kgpha, 5), p_rate2 = round_any(p_kgpha, 2), k_rate2 = round_any(k_kgpha, 10)) |> dplyr::mutate(treatment_id = 
paste(n_rate2, p_rate2, k_rate2, sep = "_")) |> dplyr::mutate(location = 
paste(NAME_2, NAME_3, sep = "_"))
dim(wheat_joined)
```
#### 4. Spatial locations of observations

Check the observations that lies inside the wheat crop mask area

```{r, message=FALSE}
eth <- geodata::gadm(path = ".", "eth", level = 0) |> st_as_sf()
class(eth)
```

Identify points that lie outside of Ethiopia
```{r}
wheat_joined2 <- wheat_joined |> terra::vect(geom = c('X', 'Y'), crs = "epsg:4326") |> st_as_sf() 
ggplot() +
  geom_sf(data = eth,
          fill = NA,
          color = "black") +
  geom_sf(data = wheat_joined2,
          aes(color = NAME_1, size = grain_yield_kgpha),
          na.rm = T) +
  coord_sf() +
  theme_bw()
```

Select those points that lie outside of the boundary and flag the observations
```{r}
flag1 <- wheat_joined |> dplyr::filter(is.na(NAME_1) | is.na(NAME_2) | is.na(NAME_3))
rmarkdown::paged_table(flag1, options=NULL)
```

Let's add a comment column for the flagged data specifying the reason for flag.

```{r}
flag1$comment <- "geolocation error"
flag1$description <- "Outside of the boundary of Ethiopia"
```

No of flagged data by treatment id
```{r}
flag1 |> dplyr::group_by(source, treatment_id) |> dplyr::summarise(n = n()) |> dplyr::arrange(desc(n))
```

Let's remove the observations outside of Ethiopia
```{r}
wheat_filtered <- wheat_joined |> dplyr::filter(!(unique_id %in% flag1$unique_id))
dim(wheat_filtered)
```

#### 5. Summarizing data

Boxplot for all data
```{r}
wheat_filtered |> ggplot(aes(x = source, y = grain_yield_kgpha, fill = source))+
  geom_boxplot()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = 'none')
```

Drop observations with yield values > 15ton/ha
```{r}
wheat_filtered <- wheat_filtered |> dplyr::filter(grain_yield_kgpha <= 15000)
```

View the summary of the numeric variables (i.e. n,p,k,grain_yield)
```{r}
for(i in c(10, 17:19)){
  message(colnames(wheat_filtered)[i])
  print(summary(wheat_filtered[,i]))
}
```
Distribution of treatments and yield
```{r}
par(mfrow=c(2,2))
hist(wheat_filtered$grain_yield_kgpha, xlab = "yield", main = "")
hist(wheat_filtered$n_rate2, xlab = "N rate", main = "")
hist(wheat_filtered$p_rate2, xlab = "P rate", main = "")
hist(wheat_filtered$k_rate2, xlab = "K rate", main = "")
```

Summarize data collected in different year
```{r  warning=FALSE, fig.width=10}
wheat_filtered |> ggplot(aes(year))+
  geom_bar(fill = "lightblue")+
  xlab("year")+
  ylab("observation")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))
```

Let's check the number of observations by source
```{r}
obs_src <- wheat_filtered |> dplyr::group_by(source) |> dplyr::summarise(n = n()) |> 
  dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_src, options = NULL) 
```

Let's check the number of observations by treatment
```{r}
obs_trt <- wheat_filtered |> dplyr::group_by(treatment_id) |> dplyr::summarise(n = n()) |> dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_trt, options=NULL) 
```

Let's check the number of observations by treatment and source
```{r}
obs_trt_src <- wheat_filtered |> dplyr::group_by(source, treatment_id) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_trt_src, options=NULL) 
```

Let's check the number of observations by source & trial id
```{r}
obs_src_trid <- wheat_filtered |> dplyr::group_by(source, trial_id) |> dplyr::summarize(n = n()) |> dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_src_trid, options=NULL) 
```

Trials with only one data points (lat, long, year and source)
```{r fig.height= 7}
few_observs <- wheat_filtered |> dplyr::group_by(trial_id) |> dplyr::summarise(n = n()) |> dplyr::filter(n == 1)
wheat_few_observs <- wheat_filtered |> dplyr::filter((trial_id %in% few_observs$trial_id))
wheat_few_observs |> ggplot(aes(x = source, y = grain_yield_kgpha, fill = source))+
  geom_point()+
  facet_wrap(~year)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8), legend.position = 'none', strip.text.x = element_text(size = 12))
```

```{r}
dim(few_observs)
dim(wheat_filtered)
```

Finding very close points to form a trial_id 
```{r}
wheat_filtered <- wheat_filtered |> dplyr::mutate(long2 = round(X, digits = 2), lat2 = round(Y , digits = 2), trial_id2 = paste(long2, lat2, year, source, sep = "_"))  
```

Dimension few observations when long lat rounded to 2 digits
```{r}
few_observs2 <- wheat_filtered |> dplyr::group_by(trial_id2) |> dplyr::summarise(n = n()) |> dplyr::filter(n == 1)
dim(few_observs2)
```

##### Dimension of number of trials with 1 data points. While merged by rounding lon & lat to 3 digits 1 data point issue will be reduced from 1213 - 1175 unique trials. But if we round to 2 digits it again reduces from 1213 to 687 unique trials. 

No of observations per region 
```{r}
wheat_filtered |> dplyr::group_by(NAME_1) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(desc(n))
```

##### Benishangul has only 5 points and may not be representative?? We suggest to generate recommendation as per agroecology level

No of observations by region and source
```{r}
obs_src_reg <- wheat_filtered |> dplyr::group_by(source, NAME_1) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_src_reg, options=NULL) 
```

No of observations by region and treatment
```{r}
obs_reg_trt <- wheat_filtered |> dplyr::group_by(treatment_id, NAME_1) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(treatment_id)
rmarkdown::paged_table(obs_reg_trt, options=NULL) 
```

Yield ranges per region
```{r, fig.width = 8}
wheat_filtered |>
ggplot(aes(y = grain_yield_kgpha, fill = NAME_1))+
  geom_boxplot(width = 0.2, outlier.colour = 'red')+
  theme_bw()+
  theme(axis.text.x = element_blank(), legend.position = "right")+
   guides(fill = guide_legend(nrow = 4))
```

Now let's see the average yield by zone
```{r}
avg_yld_zone <- wheat_filtered |> dplyr::group_by(NAME_1, NAME_2) |> dplyr::summarize(mean_yield = mean(grain_yield_kgpha)) |> dplyr::arrange(desc(mean_yield))
rmarkdown::paged_table(avg_yld_zone, options=NULL) 
```

Yield difference by zone
```{r}
wheat_filtered |> 
ggplot(aes(y = grain_yield_kgpha, fill = NAME_2))+
  geom_boxplot(width = 0.2, outlier.colour = 'red')+
  facet_wrap(~NAME_1)+
  theme_bw()+
  theme(axis.text.x = element_blank(), legend.position = "none")+
   guides(fill = guide_legend(nrow = 4))
```

Now let's plot the average yield by district
```{r}
avg_yld_dist <- wheat_filtered |> dplyr::group_by(location) |> dplyr::summarize(mean_yield = mean(grain_yield_kgpha)) |> dplyr::arrange(desc(mean_yield))
rmarkdown::paged_table(avg_yld_dist, options=NULL) 
```

Few Observations by year
```{r}
obs_few_year <- wheat_filtered |> dplyr::group_by(year) |> dplyr::summarise(n = n()) |> dplyr::arrange(n)
rmarkdown::paged_table(obs_few_year, options=NULL) 
```

Few observations above by source
Observations by year
```{r}
few_year <- wheat_filtered |> dplyr::filter(year %in% c(1992, 1990, 1991, 2008, 2009, 2010)) |> 
  dplyr::group_by(year, trial_id2) |> dplyr::summarise(n = n()) |> dplyr::arrange(n)
few_year
```

Few observations by year, source and zone
```{r}
wheat_filtered |> dplyr::filter(year %in% c(1992, 1990, 1991, 2008, 2009, 2010)) |>
  ggplot(aes(x = treatment_id, y = grain_yield_kgpha, color = source))+
  geom_point(size = 2)+
  facet_grid(NAME_2 ~ year)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```


Let's check the distribution of yield by district. 
```{r}
# Set tmap to view mode
eth_reg <- geodata::gadm(path = ".", country = "Eth", level = 1) |> st_as_sf()
wheat_filtered2 <- wheat_filtered |> 
  terra::vect(geom = c('X', 'Y'), crs = 'epsg:4326') |> st_as_sf()
tmap_mode("view")
tm_shape(eth)+
  tm_borders(col = "black") +
tm_shape(eth_reg) +
  tm_borders(col = "gray")+
tm_shape(wheat_filtered2)+
  tm_bubbles(col = "grain_yield_kgpha",
          title.col="Grain Yield (kg/ha)", 
          popup.vars=c("Zone"="NAME_2","DIstrict"="NAME_3",
                       "treatment"="treatment_id",
                       "Grain Yield"="grain_yield_kgpha", 
                       "Trial ID"="trial_id"),
          popup.format=list(grain_yield_kgpha=list(digits=2))) 
```

From the above map there is no spatial pattern of grain yield.

##### 6. Outliers

Even though the grain yield more than 7ton/ha is considered as an outlier let's drop all observations with yield values above 15ton/ha and look the distribution of the data.
```{r}
wheat_filtered <- wheat_filtered |> dplyr::filter(grain_yield_kgpha <= 15000) 
wheat_filtered |> 
  ggplot(aes(x = source, y = grain_yield_kgpha, fill = source))+
  geom_boxplot()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = 'none')
```

Let's check the above dominant outliers between 7.5ton/ha and 15ton/ha to investigate whether these are outliers or not using location, year and source (trial id)

```{r}
unq_tr_id <- wheat_filtered |> dplyr::filter(grain_yield_kgpha >= 7500 & grain_yield_kgpha <= 15000)
unq_tr_id <- unique(unq_tr_id$trial_id2)
unq_tr_id2 <- wheat_filtered |> dplyr::filter(grain_yield_kgpha <= 15000 & source %in% c("ICRISAT_ETH_Africa Rising", "ATA_Balanced_Fertilizer_Trial", "SG 2000") & year %in% c(2014, 2015, 2016) & trial_id2 %in% unq_tr_id)

 unq_tr_id2 |> 
  ggplot(aes(x = trial_id2, y = grain_yield_kgpha, fill = source))+
  geom_boxplot()+
  facet_wrap(~year, scales = "free")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8), legend.position = 'none')
```

Nutrient use for the above yield data
```{r}
col_high_yld <- wheat_filtered |> dplyr::mutate(cy = ifelse(trial_id %in% unq_tr_id2$trial_id, 'green', 'red'))
col_high_yld |> 
  ggplot(aes(x = n_rate2, y = grain_yield_kgpha, color = cy))+
  geom_point()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8), legend.position = 'none')
```

Nutrient use for the above yield data
```{r}
col_high_yld <- wheat_filtered |> dplyr::mutate(cy = ifelse( trial_id %in% unq_tr_id2$trial_id, 'green', 'red'))
col_high_yld |> 
  ggplot(aes(x = p_rate2, y = grain_yield_kgpha, color = cy))+
  geom_point()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8), legend.position = 'none')
```

For the two trials with lower yield above, we checked the treatment by yield.  
```{r}
unq_tr_id2 |> dplyr::filter(grain_yield_kgpha <= 2500 & year %in% c(2014, 2015)) |> 
  ggplot(aes(x = treatment_id, y = grain_yield_kgpha))+
  geom_point(color = "blue")+
  facet_wrap(~trial_id2)+
  theme_bw()
```

Let's see control yield distribution by region
```{r}
wheat_filtered |> dplyr::filter(treatment_id == "0_0_0") |>
ggplot(aes(y = grain_yield_kgpha, x = year, fill = as.factor(year)))+
  geom_boxplot()+
  ylab("yield")+
  facet_wrap(~NAME_1)+
  theme_bw()+
  theme(axis.text.x = element_blank(), axis.title.x =  element_blank())
```

```{r, fig.height=10, fig.width=13}
wheat_filtered |> dplyr::filter(treatment_id == "0_0_0" & year %in% c(2020,2021) & NAME_1 %in% c('Amhara', 'Oromia', 'SNNP')) |>
ggplot(aes(y = grain_yield_kgpha, x = NAME_2, fill = as.factor(year)))+
  geom_boxplot()+
  ylab("yield")+
  facet_wrap(~source, scale = 'free_x')+
  theme_bw()+
  xlab("Zone")+
  theme(axis.text.x = element_text(size = 14, angle = 15, hjust = 1), strip.text.x = element_text(size = 10))
```

First we select all trials with control, then we subset where control yield >=3.5ton/ha, we took all treatments of those trials and check the yield distribution. A box plot for those high control trials

<!-- The assumption is, the yield for other treatments must be bigger than the control yield.  -->
```{r, fig.height=10, fig.width=13}
ctrl_high_yld <- wheat_filtered |> dplyr::filter(treatment_id == "0_0_0"  & grain_yield_kgpha >= 3500 & grain_yield_kgpha <= 15000) 
wheat_filtered |> dplyr::filter(trial_id %in% unique(ctrl_high_yld$trial_id)) |>
ggplot(aes(y = grain_yield_kgpha, x = NAME_2, fill = as.factor(year)))+
  geom_boxplot()+
  geom_abline(slope = 0, intercept = 3500)+
  ylab("yield")+
  facet_wrap(~source, scale = 'free_x')+
  theme_bw()+
  xlab("Zone")+
  theme(axis.text.x = element_text(size = 14, angle = 15, hjust = 1), strip.text.x = element_text(size = 14))
```

#### 7. Yield Effect

Yield effect (from Meklit) - there are different values of the high NPK yield values.
```{r, warning=FALSE}
options(dplyr.summarise.inform = FALSE)
unique_tr_id <- unique(wheat_filtered$trial_id2)
yield_diff_mean <- data.frame()
for(i in 1:length(unique_tr_id)){
  unique_tr_yld <- wheat_filtered |> dplyr::filter(trial_id2 ==  unique_tr_id[i])
  max_n <- max(unique_tr_yld$n_rate2)
  max_p <- max(unique_tr_yld$p_rate2)
  max_k <- max(unique_tr_yld$k_rate2)
  unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p & k_rate2 == max_k)
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p)
  } 
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n)
  }
  if(nrow(unique_tr_yld2) > 1){
      ref_yld <- median(unique_tr_yld2$grain_yield_kgpha)
  }else{
      ref_yld <- unique_tr_yld2$grain_yield_kgpha
  }
unique_tr_yld <- unique_tr_yld |> dplyr::mutate(yield_diff = ref_yld - grain_yield_kgpha)
yield_diff_mean <- rbind(yield_diff_mean, unique_tr_yld)
}
saveRDS(yield_diff_mean, "yield_diff.rds")
```

Scatter plot of the yield difference
```{r}
yield_diff_mean <- yield_diff_mean |> dplyr::mutate(ref_trt = ifelse(yield_diff == 0, T, F))
ref_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff == 0) |> dplyr::select(trial_id2, yield_diff, treatment_id, grain_yield_kgpha)
other_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff != 0) |> dplyr::select(trial_id2, yield_diff, treatment_id, grain_yield_kgpha)
all_diff_data <- ref_yld_data |> dplyr::inner_join(other_yld_data, by = "trial_id2")

all_diff_data  |> ggplot(aes(y = grain_yield_kgpha.x, x = grain_yield_kgpha.y))+
  geom_point()+
  ylab("reference yield")+
  xlab("other treatements yield")+
  geom_abline(intercept = 0, slope = 1, color = 'blue')+
  xlim(0, 10000)+
  ylim(0, 10000)+
  theme_bw()
  #facet_wrap(~year)
```


#### 8. Linear mixed effects model

```{r}
#create variables to deal with scale issues:
wheat_filtered$year <- as.factor(wheat_filtered$year)
wheat_filtered$source <- as.factor(wheat_filtered$source)
wheat_filtered$location <- as.factor(wheat_filtered$location)
wheat_filtered$trial_id2 <- as.factor(wheat_filtered$trial_id2)
wheat_filtered$NAME_2 <- as.factor(wheat_filtered$NAME_2)

wheat_filtered$N100 <- wheat_filtered$n_rate2/100
wheat_filtered$P100 <- wheat_filtered$p_rate2/100
wheat_filtered$K100 <- wheat_filtered$k_rate2/100
fita <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + p_rate2 + k_rate2 + year + (1|trial_id2), data= wheat_filtered)

plot(fita)
fitb <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + p_rate2 + k_rate2 + year + (1|trial_id2) + source, data = wheat_filtered)
anova(fita, fitb)

fitc <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + p_rate2 + k_rate2 + year + NAME_2 + (1|trial_id2) + source, data = wheat_filtered)
anova(fitb, fitc)

fitd <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + p_rate2 + k_rate2 + (1|year) + NAME_2 + (1|trial_id2) + source, data = wheat_filtered)
anova(fitc, fitd)

fite <- update(fitc, . ~ . + N100:P100 + N100:K100 + P100:K100 + N100:P100:K100)
anova(fitc, fite)

fitf <- update(fite, . ~ . +(0 + N100|trial_id2) +(0 + P100|trial_id2) +(0 + K100|trial_id2))
anova(fite, fitf)
#r.squaredGLMM(fitf) 
plot(fitf)
summary(fitf)
wheat_filtered$blup <- predict(fitf, wheat_filtered)**2
```

Replace the grain yield by BLUP
```{r, warning=FALSE}
options(dplyr.summarise.inform = FALSE)
unique_tr_id <- unique(wheat_filtered$trial_id2)
yield_diff_mean <- data.frame()
for(i in 1:length(unique_tr_id)){
  unique_tr_yld <- wheat_filtered |> dplyr::filter(trial_id2 ==  unique_tr_id[i])
  max_n <- max(unique_tr_yld$n_rate2)
  max_p <- max(unique_tr_yld$p_rate2)
  max_k <- max(unique_tr_yld$k_rate2)
  unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p & k_rate2 == max_k)
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p)
  } 
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n)
  }
  if(nrow(unique_tr_yld2) > 1){
      ref_yld <- median(unique_tr_yld2$blup)
  }else{
      ref_yld <- unique_tr_yld2$blup
  }
unique_tr_yld <- unique_tr_yld |> dplyr::mutate(yield_diff = ref_yld - blup)
yield_diff_mean <- rbind(yield_diff_mean, unique_tr_yld)
}
saveRDS(yield_diff_mean, "yield_diff_blup.rds")
```

Scatter plot of the yield difference with BLUPS
```{r}
yield_diff_mean <- yield_diff_mean |> dplyr::mutate(ref_trt = ifelse(yield_diff == 0, T, F))
ref_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff == 0) |> dplyr::select(trial_id2, yield_diff, treatment_id, blup)
other_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff != 0) |> dplyr::select(trial_id2, yield_diff, treatment_id, blup)
all_diff_data <- ref_yld_data |> dplyr::inner_join(other_yld_data, by = "trial_id2")

all_diff_data  |> ggplot(aes(y = blup.x, x = blup.y))+
  geom_point()+
  ylab("reference yield")+
  xlab("other treatements yield")+
  geom_abline(intercept = 0, slope = 1, color = 'blue')+
  xlim(0, 10000)+
  ylim(0, 10000)+
  theme_bw()
  #facet_wrap(~year)
```




