---
title: "Exploratory Data Analysis - DST Harmonization Tef Data"
author: "DST team"
date: "2024-02-14"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load required packages

### This chunk loads necessary R packages for data manipulation, spatial analysis, and visualization.

```{r, message=FALSE, warning=F}
library(tidyverse)
library(sf)
library(terra)
library(klippy)
library(geodata)
library(plyr)
library(tmap)
library(tmaptools)
library(lme4)
library(simr)
```


## 2.Structure of the data 

### Reads the Tef dataset from a CSV file, specifying that the first row contains headers, and sets the file encoding to Latin1.

```{r}
Tef <- read.csv("~/shared-data/Data/Tef/fieldData/teff_dst_team_shared_v1.csv", header = T, sep = ",", fileEncoding="latin1")

nprate <- read.csv("~/shared-data/Data/Tef/fieldData/TefNPrate_July24.csv")## send by Dr Birru and Dejene
nprate <- nprate[, colnames(Tef)]

TefNPrate_AdetJuly24 <- read.csv("~/shared-data/Data/Tef/fieldData/TefNPrate_AdetJuly24.csv")## send by Dr Birhanu
nprateAdet <- TefNPrate_AdetJuly24[, colnames(Tef)]

NPrate_Beza <- read.csv("~/shared-data/Data/Tef/fieldData/NP rate data_Beza.csv")## send by Beza and Dr Birru
NPrate_Beza$treatment <- paste0("N_", NPrate_Beza$n_kgpha, "_P_", NPrate_Beza$p_kgpha)
NPrate_Beza <- NPrate_Beza[, colnames(Tef)]

Tef <- Tef %>% rbind(nprate) %>% unique()
Tef <- Tef %>% rbind(nprateAdet) %>% unique()
Tef <- Tef %>% rbind(NPrate_Beza) %>% unique()

length(unique(Tef$unique_id))
nrow(Tef)
```

### Displays the number of rows and columns in the Tef dataset.

```{r}
dim(Tef)
```

### Lists all column names of the Tef dataset.

```{r}
colnames(Tef)
```

###  Provides a concise summary of the Tef dataset, including the data type of each column.

```{r}
glimpse(Tef)
```

### Shows the first three rows of the dataset.

```{r}
head_Tef <- head(Tef,3)
rmarkdown::paged_table(head_Tef, options=NULL) 
```

### Calculates and displays the number of NA (missing) values in each column of the dataset.

```{r}
Tef_sumr <- Tef |>
  summarise_all(~sum(is.na(.)))
rmarkdown::paged_table(Tef_sumr, options=NULL) 
```

## 3. Data joining and creating combined columns

```{r}
eth_dist <- geodata::gadm(path = ".", "eth", level = 3) |> st_as_sf()
Tef_pts <- terra::vect(Tef, geom = c('long', 'lat'), crs = "epsg:4326") |> st_as_sf()
dim(Tef_pts)

Tef_joined <- st_join(Tef_pts, eth_dist) 
Tef_joined <- st_coordinates(Tef_joined) |> cbind(as.data.frame(Tef_joined))  |> dplyr::select(unique_id, X, Y, year, source, treatment, n_kgpha, p_kgpha, k_kgpha, grain_yield_kgpha, NAME_1, NAME_2, NAME_3)
head(Tef_joined, 3)
```

Let's create a trial id by concatenating long, lat, year and source. Besides we round the treatment values and concatenate to create a unique treatment id.
before that, because there are treatments from the same source, same time and very close locations having different GPS, (4 digits only on lat), 
we choose to round the GPS to bring eh different treatments from the same trial have one locations 

Finding very close points to form a trial_id 
```{r}
Tef_joined <- Tef_joined |> dplyr::mutate(long2 = round(X, digits = 3), lat2 = round(Y , digits = 3), trial_id = paste(long2, lat2, year, source, sep = "_"),trial_id2 = paste(X, Y, year, source, sep = "_"))  

```

## 4. sorting out the issue between unique ID and trial ID
There are trial IDs with different unique ID, and this is causing duplication in the data
identify those data below and drop the unique_id and use trial ID from here on

```{r, unique ID verus trial ID}
tids_moreids <- Tef_joined %>% 
  dplyr::group_by(trial_id, unique_id) %>%  
  dplyr::summarise(n = n()) 
tids_moreids[tids_moreids$n > 1, ]

Tef_joined1 <- Tef_joined[Tef_joined$trial_id %in% tids_moreids$trial_id, ] 
dim(Tef_joined)

Tef_joined <- Tef_joined %>% dplyr::select(-c(unique_id)) %>% unique()
dim(Tef_joined)

length(unique(Tef_joined$trial_id))
length(unique(Tef_joined$trial_id2))
```

```{r}
Tef_joined <- Tef_joined |> dplyr::mutate(NAME_1 = replace(NAME_1, NAME_1 == "Southern Nations, Nationalities", "SNNP"))|> dplyr::mutate(n_rate2 = round_any(n_kgpha, 5), p_rate2 = round_any(p_kgpha, 5), k_rate2 = round_any(k_kgpha, 10)) |> dplyr::mutate(treatment_id = 
paste(n_rate2, p_rate2, k_rate2, sep = "_")) |> dplyr::mutate(location = 
paste(NAME_2, NAME_3, sep = "_"))
dim(Tef_joined)
```
## 5. remove points with yield > 5000 kg/ha and < 300 kg/ha, removes 192 data points

```{r}
unique(Tef_joined[Tef_joined$grain_yield_kgpha < 300, ]$treatment_id)

dim(Tef_joined %>% dplyr::filter(grain_yield_kgpha > 5000, ))
dim(Tef_joined %>% dplyr::filter(grain_yield_kgpha < 300, ))
Tef_joined <- Tef_joined %>% dplyr::filter(grain_yield_kgpha > 300, )
Tef_joined <- Tef_joined %>% dplyr::filter(grain_yield_kgpha <= 5000, )

```


## 6. Spatial locations of observations

Check the observations that lies inside the Tef crop mask area

```{r, message=FALSE}
eth <- geodata::gadm(path = ".", "eth", level = 0) |> st_as_sf()
class(eth)
```

Identify points that lie outside of Ethiopia
```{r}

Tef_joined <- Tef_joined %>% dplyr::filter(long2 < 50) %>% unique()
Tef_joined2 <- Tef_joined |> terra::vect(geom = c('X', 'Y'), crs = "epsg:4326") |> st_as_sf() 
ggplot() +
  geom_sf(data = eth,
          fill = NA,
          color = "black") +
  geom_sf(data = Tef_joined2,
          aes(color = NAME_1, size = grain_yield_kgpha),
          na.rm = T) +
  coord_sf() +
  theme_bw()
```

Select those points that lie outside of the boundary and flag the observations
```{r}
flag1 <- Tef_joined |> dplyr::filter(is.na(NAME_1) | is.na(NAME_2) | is.na(NAME_3))
rmarkdown::paged_table(flag1, options=NULL)
```

Let's add a comment column for the flagged data specifying the reason for flag.

```{r}
flag1$comment <- "geolocation error"
flag1$description <- "Outside of the boundary of Ethiopia"
```

No of flagged data by treatment id
```{r}
flag1 |> dplyr::group_by(source, treatment_id) |> dplyr::summarise(n = n()) |> dplyr::arrange(desc(n))
```

Let's remove the observations outside of Ethiopia
```{r}
Tef_filtered <- Tef_joined |> dplyr::filter(!(trial_id %in% flag1$trial_id))
dim(Tef_filtered)
```

## 7. Number of observations by region and global yield data distibution 
No of observations per region 
```{r}
Tef_filtered |> dplyr::group_by(NAME_1) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(desc(n))
```

```{r}
Tef_filtered |> ggplot(aes(x = source, y = grain_yield_kgpha, fill = source))+
  geom_boxplot()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = 'none')
```
## 8. Idenitfy observations with yield values > 5 ton/ha and SG 2000
```{r}
dim(Tef_filtered)
sg_2000 <- Tef_filtered |> dplyr::filter(source == "SG 2000")
dim(sg_2000)

yld_gr_7k <- Tef_filtered |> dplyr::filter(grain_yield_kgpha > 5000)
Tef_filtered <- Tef_filtered |> dplyr::filter(source != "SG 2000" & grain_yield_kgpha <= 5000 & p_rate2 < 50)
dim(yld_gr_7k)
dim(Tef_filtered)
```
## 9. View the summary of n,p,k,grain_yield

```{r}
for(i in c(6:9)){
  message(colnames(Tef_filtered)[i])
  print(summary(Tef_filtered[,i]))
}
```
Distribution of treatments and yield
```{r}
par(mfrow=c(2,2))
hist(Tef_filtered$grain_yield_kgpha, xlab = "yield", main = "")
hist(Tef_filtered$n_rate2, xlab = "N rate", main = "")
hist(Tef_filtered$p_rate2, xlab = "P rate", main = "")
hist(Tef_filtered$k_rate2, xlab = "K rate", main = "")
```
## 10. Number of observations by year
Summarize data collected in different year
```{r  warning=FALSE, fig.width=10}
Tef_filtered |> ggplot(aes(year))+
  geom_bar(fill = "lightblue")+
  xlab("year")+
  ylab("observation")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))
```

## 11. Number of observations by data source

```{r}
obs_src <- Tef_filtered |> dplyr::group_by(source) |> dplyr::summarise(n = n()) |> 
  dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_src, options = NULL) 
```
## 12. Number of observations by data treatment

```{r}
obs_trt <- Tef_filtered |> dplyr::group_by(treatment_id) |> dplyr::summarise(n = n()) |> dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_trt, options=NULL) 
```

## 13. Number of observations by data treatment and source
```{r}
obs_trt_src <- Tef_filtered |> dplyr::group_by(source, treatment_id) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_trt_src, options=NULL) 
```


## 14. Number of observations by trial ID
```{r}
obs_src_trid <- Tef_filtered |> dplyr::group_by(source, trial_id) |> dplyr::summarize(n = n()) |> dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_src_trid, options=NULL) 
```


## 15. Trials with only one data points (lat, long, year and source)
```{r fig.height= 7}
few_observs <- Tef_filtered |> dplyr::group_by(trial_id) |> dplyr::summarise(n = n()) |> dplyr::filter(n == 1)
Tef_few_observs <- Tef_filtered |> dplyr::filter((trial_id %in% few_observs$trial_id))
Tef_few_observs |> ggplot(aes(x = source, y = grain_yield_kgpha, fill = source))+
  geom_point()+
  facet_wrap(~year)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8), legend.position = 'none', strip.text.x = element_text(size = 12))
```

```{r}
dim(Tef_few_observs)
dim(Tef_filtered)
```


## 16. Excluding dat flaggged to be checked 
```{r warning=FALSE}
few_obs_data <- Tef_filtered |> dplyr::filter(trial_id %in% few_observs$trial_id)

few_obs_data$comment <- "only one observation per trial"
# yld_gr_7k$comment <- "greater than 5000 kg/ha"
sg_2000$comment <- "SG 2000 data" 
flagged_data <- rbind(sg_2000, few_obs_data)
dim(flagged_data)

#write.csv(flagged_data, "./EIAR_toCheck/flagged_data.csv", col.names = T, row.names = F)
```

```{r}
### the SG 2000 and yield > 5 t/ha are already removed above
Tef_filtered <- Tef_filtered |> dplyr::filter(!(trial_id %in% flagged_data$trial_id))
dim(Tef_filtered)
```


## 17. ISFM treatments
 407 trial IDs, from these we are excluding the treatments with organic addition: this drops ~ 44 points
```{r}
Tef_filtered$index <- c(1:nrow(Tef_filtered))
orgIndex <- grep("Org", Tef_filtered$treatment)
orgdata <- droplevels(Tef_filtered[Tef_filtered$index %in% orgIndex, ])
Tef_filtered <- droplevels(Tef_filtered[!Tef_filtered$index %in% orgIndex, ])
dim(Tef_filtered)
```
## 18. same yield for several tereatments within a trial
Let's check the trials that have the same yield for different treatments in a trial. It seems a rep mean yield is 
assigned to all treatments in rep. 

```{r}
Tef_filtered$index <- paste(Tef_filtered$trial_id, Tef_filtered$treatment_id, sep="_")
dd <- NULL
for (tids in unique(Tef_filtered$trial_id)){
  tdata <- Tef_filtered[Tef_filtered$trial_id == tids, ]
  ttable <- as.data.frame(table(tdata$grain_yield_kgpha))
  ttable <- droplevels(ttable[ttable$Freq > 1,])
  if(nrow(ttable) > 0){
     tdata <- tdata[tdata$grain_yield_kgpha %in% ttable$Var1, ]
  dd <- rbind(dd, tdata)
  }
}
dd <- dd %>% dplyr::select(c(index, trial_id)) %>% unique()
dim(dd)
length(unique(dd$trial_id))
```

Let's filter out the above observations from the Tef filtered data
```{r}
dim(Tef_filtered)
length(unique(Tef_filtered$trial_id))
Tef_filtered <- Tef_filtered %>% dplyr::filter(!index %in% dd$index) %>% unique()
dim(Tef_filtered)
length(unique(Tef_filtered$trial_id))
```

```{r}
obs_src_reg <- Tef_filtered |> dplyr::group_by(source, NAME_1) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_src_reg, options=NULL) 
```

No of observations by region and treatment
```{r}
obs_reg_trt <- Tef_filtered |> dplyr::group_by(treatment_id, NAME_1) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(treatment_id)
rmarkdown::paged_table(obs_reg_trt, options=NULL) 
```
## 19. Yield ranges per region

```{r, fig.width = 8}
Tef_filtered |>
ggplot(aes(y = grain_yield_kgpha, fill = NAME_1))+
  geom_boxplot(width = 0.2, outlier.colour = 'red')+
  theme_bw()+
  theme(axis.text.x = element_blank(), legend.position = "right")+
   guides(fill = guide_legend(nrow = 4))
```


## 20. Yield ranges per Zone and region
```{r}
avg_yld_zone <- Tef_filtered |> dplyr::group_by(NAME_1, NAME_2) |> dplyr::summarize(mean_yield = mean(grain_yield_kgpha)) |> dplyr::arrange(desc(mean_yield))
rmarkdown::paged_table(avg_yld_zone, options=NULL) 
```

Within and between zone and region Yield difference
```{r}
Tef_filtered |> 
ggplot(aes(y = grain_yield_kgpha, fill = NAME_2))+
  geom_boxplot(width = 0.2, outlier.colour = 'red')+
  facet_wrap(~NAME_1)+
  theme_bw()+
  theme(axis.text.x = element_blank(), legend.position = "none")+
   guides(fill = guide_legend(nrow = 4))
```

The average yield by district
```{r}
avg_yld_dist <- Tef_filtered |> dplyr::group_by(location) |> dplyr::summarize(mean_yield = mean(grain_yield_kgpha)) |> dplyr::arrange(desc(mean_yield))
rmarkdown::paged_table(avg_yld_dist, options=NULL) 
```

Few Observations by year
```{r}
obs_few_year <- Tef_filtered |> dplyr::group_by(year) |> dplyr::summarise(n = n()) |> dplyr::arrange(n)
rmarkdown::paged_table(obs_few_year, options=NULL) 
```

Observations by year
```{r}
few_year <- Tef_filtered |> dplyr::filter(year %in% c(1992, 1990, 1991, 2008, 2009, 2010, 2017)) |> 
  dplyr::group_by(year, trial_id) |> dplyr::summarise(n = n()) |> dplyr::arrange(n)
few_year
```

The data for years with observations are sourced from literature, isolated students work for PhD or MSc
```{r}
Tef_filtered |> dplyr::filter(year %in% c(1992, 1990, 1991, 2008, 2009, 2010)) |>
  ggplot(aes(x = treatment_id, y = grain_yield_kgpha, color = source))+
  geom_point(size = 2)+
  facet_grid(NAME_2 ~ year)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

## 21. Interactive map for yield  
```{r}
# Set tmap to view mode
eth_reg <- geodata::gadm(path = ".", country = "Eth", level = 1) |> st_as_sf()
Tef_filtered2 <- Tef_filtered |> 
  terra::vect(geom = c('X', 'Y'), crs = 'epsg:4326') |> st_as_sf()
tmap_mode("view")
tm_shape(eth)+
  tm_borders(col = "black") +
tm_shape(eth_reg) +
  tm_borders(col = "gray")+
tm_shape(Tef_filtered2)+
  tm_bubbles(col = "grain_yield_kgpha",
          title.col="Grain Yield (kg/ha)", 
          popup.vars=c("Zone"="NAME_2","DIstrict"="NAME_3",
                       "treatment"="treatment_id",
                       "Grain Yield"="grain_yield_kgpha", 
                       "Trial ID"="trial_id"),
          popup.format=list(grain_yield_kgpha=list(digits=2))) 
```

From the above map there is no spatial pattern of grain yield.



## 22. final selected data

```{r}
Tef_filtered <- Tef_filtered |> dplyr::filter(grain_yield_kgpha <= 15000) 
Tef_filtered |> 
  ggplot(aes(x = source, y = grain_yield_kgpha, fill = source))+
  geom_boxplot()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = 'none')
```


## 23. Yield from the 0:0:0 treatment, the control by Zone and Region
It can help to understand regions with high baseline yield
```{r}
Tef_filtered |> dplyr::filter(treatment_id == "0_0_0") |>
ggplot(aes(y = grain_yield_kgpha, x = year, fill = as.factor(year)))+
  geom_boxplot()+
  ylab("yield")+
  facet_wrap(~NAME_1)+
  theme_bw()+
  theme(axis.text.x = element_blank(), axis.title.x =  element_blank())
```

```{r, fig.height=10, fig.width=13}
Tef_filtered |> dplyr::filter(treatment_id == "0_0_0" & year %in% c(2020,2021) & NAME_1 %in% c('Amhara', 'Oromia', 'SNNP')) |>
ggplot(aes(y = grain_yield_kgpha, x = NAME_2, fill = as.factor(year)))+
  geom_boxplot()+
  ylab("yield")+
  facet_wrap(~source, scale = 'free_x')+
  theme_bw()+
  xlab("Zone")+
  theme(axis.text.x = element_text(size = 14, angle = 15, hjust = 1), strip.text.x = element_text(size = 10))
```

## 24. Yield Effect

Reference treatment need to be selected within every trial as there are different NPK rates tested.
```{r, warning=FALSE}
options(dplyr.summarise.inform = FALSE)
unique_tr_id <- unique(Tef_filtered$trial_id)
yield_diff_mean <- data.frame()
for(i in 1:length(unique_tr_id)){
  unique_tr_yld <- Tef_filtered |> dplyr::filter(trial_id ==  unique_tr_id[i])
  max_n <- max(unique_tr_yld$n_rate2)
  max_p <- max(unique_tr_yld$p_rate2)
  max_k <- max(unique_tr_yld$k_rate2)
  unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p & k_rate2 == max_k)
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p)
  } 
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n)
  }
  if(nrow(unique_tr_yld2) > 1){
      ref_yld <- median(unique_tr_yld2$grain_yield_kgpha)
  }else{
      ref_yld <- unique_tr_yld2$grain_yield_kgpha
  }
unique_tr_yld <- unique_tr_yld |> dplyr::mutate(yield_diff = ref_yld - grain_yield_kgpha)
yield_diff_mean <- rbind(yield_diff_mean, unique_tr_yld)
}
saveRDS(yield_diff_mean, "~/shared-data/Data/Tef/fieldData/yield_diff.rds")
```

Scatter plot of the yield difference: there are quite some negative yield effects in the data
```{r}
yield_diff_mean <- yield_diff_mean |> dplyr::mutate(ref_trt = ifelse(yield_diff == 0, T, F))
ref_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff == 0) |> dplyr::select(trial_id, yield_diff, treatment_id, grain_yield_kgpha)
other_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff != 0) |> dplyr::select(trial_id, yield_diff, treatment_id, grain_yield_kgpha)
all_diff_data <- ref_yld_data |> dplyr::inner_join(other_yld_data, by = "trial_id")

all_diff_data  |> ggplot(aes(y = grain_yield_kgpha.x, x = grain_yield_kgpha.y))+
  geom_point()+
  ylab("reference yield")+
  xlab("other treatements yield")+
  geom_abline(intercept = 0, slope = 1, color = 'blue')+
  xlim(0, 5000)+
  ylim(0, 5000)+
  theme_bw()
  #facet_wrap(~year)
```

## 25. Linear mixed effects model to reduce the ranomd noise in the data

```{r}
#create variables to deal with scale issues:
Tef_filtered$year <- as.factor(Tef_filtered$year)
Tef_filtered$source <- as.factor(Tef_filtered$source)
Tef_filtered$location <- as.factor(Tef_filtered$location)
Tef_filtered$trial_id <- as.factor(Tef_filtered$trial_id)
Tef_filtered$NAME_2 <- as.factor(Tef_filtered$NAME_2)

Tef_filtered$N100 <- Tef_filtered$n_rate2/100
Tef_filtered$P100 <- Tef_filtered$p_rate2/100
Tef_filtered$K100 <- Tef_filtered$k_rate2/100

fita <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + n_rate2^2 + p_rate2 + k_rate2 + year + (1|trial_id), data= Tef_filtered)

fitb <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + n_rate2^2 + p_rate2 + k_rate2 + year + (1|trial_id) + source, data = Tef_filtered)
anova(fita, fitb)

fitc <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + n_rate2^2 + p_rate2 + k_rate2 + year + NAME_2 + (1|trial_id) + source, data = Tef_filtered)
anova(fitb, fitc)

fitd <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + n_rate2^2 + p_rate2 + k_rate2 + (1|year) + NAME_2 + (1|trial_id) + source, data = Tef_filtered)
anova(fitc, fitd)

fite <- update(fitc, . ~ . + N100:P100 + N100:K100 + P100:K100 + N100:P100:K100)
anova(fitc, fite)

fitf <- update(fite, . ~ . +(0 + N100|trial_id) +(0 + P100|trial_id) +(0 + K100|trial_id))
anova(fite, fitf)
#r.squaredGLMM(fitf) 

## fitting the model with RMLE
fitg <- update(fite, . ~ . +(0 + N100|trial_id) +(0 + P100|trial_id) +(0 + K100|trial_id))
Tef_filtered$blup <- predict(fitg, Tef_filtered)**2
```


## 26. compute the yield effect using the BLUP from the linear mixed effects model
```{r, warning=FALSE}
options(dplyr.summarise.inform = FALSE)
unique_tr_id <- unique(Tef_filtered$trial_id)
yield_diff_mean <- data.frame()
for(i in 1:length(unique_tr_id)){
  unique_tr_yld <- Tef_filtered |> dplyr::filter(trial_id ==  unique_tr_id[i])
  max_n <- max(unique_tr_yld$n_rate2)
  max_p <- max(unique_tr_yld$p_rate2)
  max_k <- max(unique_tr_yld$k_rate2)
  unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p & k_rate2 == max_k)
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p)
  } 
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n)
  }
  if(nrow(unique_tr_yld2) > 1){
      ref_yld <- median(unique_tr_yld2$blup)
  }else{
      ref_yld <- unique_tr_yld2$blup
  }
unique_tr_yld <- unique_tr_yld |> dplyr::mutate(yield_diff = ref_yld - blup)
yield_diff_mean <- rbind(yield_diff_mean, unique_tr_yld)
}

```

Scatter plot of the yield difference with BLUPS
```{r}
yield_diff_mean <- yield_diff_mean |> dplyr::mutate(ref_trt = ifelse(yield_diff == 0, T, F))
ref_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff == 0) |> dplyr::select(trial_id, yield_diff, treatment_id, blup)
other_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff != 0) |> dplyr::select(trial_id, yield_diff, treatment_id, blup)
all_diff_data <- ref_yld_data |> dplyr::inner_join(other_yld_data, by = "trial_id")

yield_diff_mean <- unique(yield_diff_mean[, c("trial_id","source", "year","NAME_1","NAME_2","NAME_3", "long2", "lat2","n_rate2", "p_rate2", "k_rate2", "treatment_id","ref_trt","grain_yield_kgpha","blup", "yield_diff")])

saveRDS(yield_diff_mean, "~/shared-data/Data/Tef/fieldData/tef_modelReady_NPrateAdded_Beza.rds")



all_diff_data  |> ggplot(aes(y = blup.x, x = blup.y))+
  geom_point()+
  ylab("reference yield")+
  xlab("other treatements yield")+
  geom_abline(intercept = 0, slope = 1, color = 'blue')+
  xlim(0, 5000)+
  ylim(0, 5000)+
  theme_bw()
  #facet_wrap(~year)
```


## 27. Checking the data distribution for the NP rate data
```{r, warning=FALSE}
plotdf <- yield_diff_mean %>% dplyr::select(-c(ref_trt, grain_yield_kgpha, yield_diff)) %>% unique()

gga <- ggplot(plotdf[plotdf$source == "NP_rate" & plotdf$year == 2023, ], aes(n_rate2, blup, col=factor(p_rate2))) +
  geom_point()+
  facet_wrap(~NAME_3)

ggb <- ggplot(plotdf[plotdf$source == "NP_rate" & plotdf$year == 2023, ], aes(n_rate2, blup, col=factor(NAME_2), group=factor(trial_id))) +
  geom_line()+
   geom_point()+
  facet_wrap(~p_rate2, scales = "free_x")


ggc <- ggplot(plotdf[plotdf$source == "NP_rate" & plotdf$year == 2023, ], aes(n_rate2, blup, group=factor(trial_id))) +
    geom_line()+
    geom_point()+
    facet_grid(NAME_2~p_rate2, scales = "free_x")


ggd <- ggplot(plotdf[plotdf$source == "NP_rate" & plotdf$year == 2023, ], aes(p_rate2, blup, col=factor(n_rate2))) +
  geom_point()+
  facet_wrap(~NAME_3)


gge <- ggplot(plotdf[plotdf$source == "NP_rate" & plotdf$year == 2023, ], aes(p_rate2, blup, col=factor(NAME_2), group=factor(trial_id))) +
  geom_line()+
   geom_point()+
  facet_wrap(~n_rate2, scales = "free_x")


ggf <- ggplot(plotdf[plotdf$source == "NP_rate" & plotdf$year == 2023, ], aes(p_rate2, blup, group=factor(trial_id))) +
    geom_line()+
    geom_point()+
    facet_grid(NAME_2~n_rate2, scales = "free_x")
# 
# suppressWarnings(print(gga))
# suppressWarnings(print(ggb))
# suppressWarnings(print(ggc))
# suppressWarnings(print(ggd))
# suppressWarnings(print(gge))
# suppressWarnings(print(ggf))


ggsave("~/shared-data/Data/Tef/fieldData/Nresponse_byZone_rawdata.pdf", gga, width=6, height = 6)
ggsave("~/shared-data/Data/Tef/fieldData/Nresponse_byP_rawdata.pdf", ggb, width=6, height = 6)
suppressWarnings(ggsave("~/shared-data/Data/Tef/fieldData/Nresponse_byP_byZone_rawdata.pdf", ggc, width=6, height = 14))
ggsave("~/shared-data/Data/Tef/fieldData/Presponse_byZone_rawdata.pdf", ggd, width=6, height = 6)
ggsave("~/shared-data/Data/Tef/fieldData/Presponse_byN_rawdata.pdf", gge, width=6, height = 6)
suppressWarnings(ggsave("~/shared-data/Data/Tef/fieldData/Presponse_byN_byZone_rawdata.pdf", ggf, width=8, height = 14))


```
