---
title: "Wheat EDA"
output: html_document
author: "The Allinace of CIAT and Bioversity"
date: "2024-02-14"
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load required packages

```{r, message=FALSE, warning=F}
library(tidyverse)
library(sf)
library(terra)
library(klippy)
library(geodata)
library(plyr)
library(tmap)
library(tmaptools)
library(lme4)
library(simr)
```


## 2.Structure of the data 

```{r}

wheat <- read.csv("~/shared-data/Data/Wheat/fieldData/wheat_dst_team_shared_v3.csv", header = T, sep = ",", fileEncoding="latin1")
length(unique(wheat$unique_id))
```

See the dimensions of the data
```{r}
dim(wheat)
```

View columns of the data
```{r}
colnames(wheat)
```

See the structure of the data
```{r}
glimpse(wheat)
```

See some observations of the whole dataset
```{r}
hed_wheat <- head(wheat,3)
rmarkdown::paged_table(hed_wheat, options=NULL) 
```

Let's check the number of NA values in each column
```{r}
wheat_sumr <- wheat |>
  summarise_all(~sum(is.na(.)))
rmarkdown::paged_table(wheat_sumr, options=NULL) 
```

From the NA values summary table we have only (17782 - 13709) values of planting and harvesting date observations

## 3. Data joining and creating combined columns


```{r}
eth_dist <- geodata::gadm(path = ".", "eth", level = 3) |> st_as_sf()
wheat_pts <- terra::vect(wheat, geom = c('long', 'lat'), crs = "epsg:4326") |> st_as_sf()
dim(wheat_pts)
wheat_joined <- st_join(wheat_pts, eth_dist) 
wheat_joined <- st_coordinates(wheat_joined) |> cbind(as.data.frame(wheat_joined))  |> dplyr::select(unique_id, X, Y, year, source, treatment, n_kgpha, p_kgpha, k_kgpha, grain_yield_kgpha, pl_date_ec, hv_date_ec, NAME_1, NAME_2, NAME_3)
head(wheat_joined, 3)
```

Let's create a trial id by concatenating long, lat, year and source. Besides we round the treatment values and concatenate to create a unique treatment id.
before that, because there are treatments from the same source, same time and very close locations having different GPS, (4 digits only on lat), 
we choose to round the GPS to bring eh different treatments from the same trial have one locations 

Finding very close points to form a trial_id 
```{r}
wheat_joined <- wheat_joined |> dplyr::mutate(long2 = round(X, digits = 2), lat2 = round(Y , digits = 2), trial_id = paste(long2, lat2, year, source, sep = "_"),trial_id2 = paste(X, Y, year, source, sep = "_"))  

```

## 4. sorting out the issue between unique ID and trial ID
There are trial IDs with different unique ID, and this is causing duplication in the data
identify those data below and drop the unique_id and use trial ID from here on

```{r, unique ID verus trial ID}
tids_moreids <- wheat_joined %>% 
  dplyr::group_by(trial_id, unique_id) %>%  
  dplyr::summarise(n = n()) 
tids_moreids[tids_moreids$n > 1, ]

wheat_joined1 <- wheat_joined[wheat_joined$trial_id %in% tids_moreids$trial_id, ] 
dim(wheat_joined)
wheat_joined <- wheat_joined %>% dplyr::select(-c(unique_id)) %>% unique()
dim(wheat_joined)
length(unique(wheat_joined$trial_id))
length(unique(wheat_joined$trial_id2))
```

```{r}
wheat_joined <- wheat_joined |> dplyr::mutate(NAME_1 = replace(NAME_1, NAME_1 == "Southern Nations, Nationalities", "SNNP"))|> dplyr::mutate(n_rate2 = round_any(n_kgpha, 5), p_rate2 = round_any(p_kgpha, 2), k_rate2 = round_any(k_kgpha, 10)) |> dplyr::mutate(treatment_id = 
paste(n_rate2, p_rate2, k_rate2, sep = "_")) |> dplyr::mutate(location = 
paste(NAME_2, NAME_3, sep = "_"))
dim(wheat_joined)
```
## 5. remove points with < 500 kg/ha, removes 192 data points

```{r}
unique(wheat_joined[wheat_joined$grain_yield_kgpha < 500, ]$treatment_id)

dim(wheat_joined %>% dplyr::filter(grain_yield_kgpha < 500, ))
wheat_joined <- wheat_joined %>% dplyr::filter(grain_yield_kgpha > 500, )

```


## 6. Spatial locations of observations

Check the observations that lies inside the wheat crop mask area

```{r, message=FALSE}
eth <- geodata::gadm(path = ".", "eth", level = 0) |> st_as_sf()
class(eth)
```

Identify points that lie outside of Ethiopia
```{r}
wheat_joined2 <- wheat_joined |> terra::vect(geom = c('X', 'Y'), crs = "epsg:4326") |> st_as_sf() 
ggplot() +
  geom_sf(data = eth,
          fill = NA,
          color = "black") +
  geom_sf(data = wheat_joined2,
          aes(color = NAME_1, size = grain_yield_kgpha),
          na.rm = T) +
  coord_sf() +
  theme_bw()
```

Select those points that lie outside of the boundary and flag the observations
```{r}
flag1 <- wheat_joined |> dplyr::filter(is.na(NAME_1) | is.na(NAME_2) | is.na(NAME_3))
rmarkdown::paged_table(flag1, options=NULL)
```

Let's add a comment column for the flagged data specifying the reason for flag.

```{r}
flag1$comment <- "geolocation error"
flag1$description <- "Outside of the boundary of Ethiopia"
```

No of flagged data by treatment id
```{r}
flag1 |> dplyr::group_by(source, treatment_id) |> dplyr::summarise(n = n()) |> dplyr::arrange(desc(n))
```

Let's remove the observations outside of Ethiopia
```{r}
wheat_filtered <- wheat_joined |> dplyr::filter(!(trial_id %in% flag1$trial_id))
dim(wheat_filtered)
```

## 7. Number of observations by region and global yield data distibution 
No of observations per region 
```{r}
wheat_filtered |> dplyr::group_by(NAME_1) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(desc(n))
```

```{r}
wheat_filtered |> ggplot(aes(x = source, y = grain_yield_kgpha, fill = source))+
  geom_boxplot()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = 'none')
```
## 8. Idenitfy observations with yield values > 7.5 ton/ha  and SG 2000 data 
```{r}
sg_2000 <- wheat_filtered |> dplyr::filter(source == "SG 2000")
dim(sg_2000)
yld_gr_7k <- wheat_filtered |> dplyr::filter(grain_yield_kgpha > 7500)
wheat_filtered <- wheat_filtered |> dplyr::filter(source != "SG 2000" & grain_yield_kgpha <= 7500)
dim(yld_gr_7k)
dim(wheat_filtered)
```
## 9. View the summary of n,p,k,grain_yield

```{r}
for(i in c(9, 19:21)){
  message(colnames(wheat_filtered)[i])
  print(summary(wheat_filtered[,i]))
}
```
Distribution of treatments and yield
```{r}
par(mfrow=c(2,2))
hist(wheat_filtered$grain_yield_kgpha, xlab = "yield", main = "")
hist(wheat_filtered$n_rate2, xlab = "N rate", main = "")
hist(wheat_filtered$p_rate2, xlab = "P rate", main = "")
hist(wheat_filtered$k_rate2, xlab = "K rate", main = "")
```
## 10. Number of observations by year
Summarize data collected in different year
```{r  warning=FALSE, fig.width=10}
wheat_filtered |> ggplot(aes(year))+
  geom_bar(fill = "lightblue")+
  xlab("year")+
  ylab("observation")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))
```

## 11. Number of observations by data source

```{r}
obs_src <- wheat_filtered |> dplyr::group_by(source) |> dplyr::summarise(n = n()) |> 
  dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_src, options = NULL) 
```
## 12. Number of observations by data treatment

```{r}
obs_trt <- wheat_filtered |> dplyr::group_by(treatment_id) |> dplyr::summarise(n = n()) |> dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_trt, options=NULL) 
```

## 13. Number of observations by data treatment and source
```{r}
obs_trt_src <- wheat_filtered |> dplyr::group_by(source, treatment_id) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_trt_src, options=NULL) 
```


## 14. Number of observations by trial ID
```{r}
obs_src_trid <- wheat_filtered |> dplyr::group_by(source, trial_id) |> dplyr::summarize(n = n()) |> dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_src_trid, options=NULL) 
```


## 15. Trials with only one data points (lat, long, year and source)
```{r fig.height= 7}
few_observs <- wheat_filtered |> dplyr::group_by(trial_id) |> dplyr::summarise(n = n()) |> dplyr::filter(n == 1)
wheat_few_observs <- wheat_filtered |> dplyr::filter((trial_id %in% few_observs$trial_id))
wheat_few_observs |> ggplot(aes(x = source, y = grain_yield_kgpha, fill = source))+
  geom_point()+
  facet_wrap(~year)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8), legend.position = 'none', strip.text.x = element_text(size = 12))
```

```{r}
dim(wheat_few_observs)
dim(wheat_filtered)
```


## 16. Excluding dat flaggged to be checked 
```{r warning=FALSE}
few_obs_data <- wheat_filtered |> dplyr::filter(trial_id %in% few_observs$trial_id)

few_obs_data$comment <- "only one observation per trial"
yld_gr_7k$comment <- "greater than 7500kg/ha"
sg_2000$comment <- "SG 2000 data" 
flagged_data <- rbind(sg_2000, yld_gr_7k, few_obs_data)
dim(flagged_data)

#write.csv(flagged_data, "./EIAR_toCheck/flagged_data.csv", col.names = T, row.names = F)
```

```{r}
wheat_filtered <- wheat_filtered |> dplyr::filter(!(trial_id %in% flagged_data$trial_id))
dim(wheat_filtered)
```


## 17. ISFM treatments
 407 trial IDs, from these we are excluding the treatments with organic addition: this drops ~ 195 points
```{r}
wheat_filtered$index <- c(1:nrow(wheat_filtered))
orgIndex <- grep("Org", wheat_filtered$treatment)
orgdata <- droplevels(wheat_filtered[wheat_filtered$index %in% orgIndex, ])
wheat_filtered <- droplevels(wheat_filtered[!wheat_filtered$index %in% orgIndex, ])
dim(wheat_filtered)
```
## 18. same yield for several tereatments within a trial
Let's check the trials that have the same yield for different treatments in a trial. It seems a rep mean yield is 
assigned to all treatments in rep. 

```{r}
wheat_filtered$index <- paste(wheat_filtered$trial_id, wheat_filtered$treatment_id, sep="_")
dd <- NULL
for (tids in unique(wheat_filtered$trial_id)){
  tdata <- wheat_filtered[wheat_filtered$trial_id == tids, ]
  ttable <- as.data.frame(table(tdata$grain_yield_kgpha))
  ttable <- droplevels(ttable[ttable$Freq > 1,])
  if(nrow(ttable) > 0){
     tdata <- tdata[tdata$grain_yield_kgpha %in% ttable$Var1, ]
  dd <- rbind(dd, tdata)
  }
}
dd <- dd %>% dplyr::select(c(index, trial_id))
dim(dd)
length(unique(dd$trial_id))
```

Let's filter out the above observations from the wheat filtered data
```{r}
dim(wheat_filtered)
length(unique(wheat_filtered$trial_id))
wheat_filtered <- wheat_filtered %>% dplyr::filter(!index %in% dd$index) %>% unique()
dim(wheat_filtered)
length(unique(wheat_filtered$trial_id))
```

Benishangul has only 5 points and may not be representative?? We suggest to generate recommendation as per agroecology level
No of observations by region and source
```{r}
obs_src_reg <- wheat_filtered |> dplyr::group_by(source, NAME_1) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(desc(n))
rmarkdown::paged_table(obs_src_reg, options=NULL) 
```

No of observations by region and treatment
```{r}
obs_reg_trt <- wheat_filtered |> dplyr::group_by(treatment_id, NAME_1) |> dplyr::summarize(n = n()) |>
  dplyr::arrange(treatment_id)
rmarkdown::paged_table(obs_reg_trt, options=NULL) 
```
## 19. Yield ranges per region

```{r, fig.width = 8}
wheat_filtered |>
ggplot(aes(y = grain_yield_kgpha, fill = NAME_1))+
  geom_boxplot(width = 0.2, outlier.colour = 'red')+
  theme_bw()+
  theme(axis.text.x = element_blank(), legend.position = "right")+
   guides(fill = guide_legend(nrow = 4))
```


## 20. Yield ranges per Zone and region
```{r}
avg_yld_zone <- wheat_filtered |> dplyr::group_by(NAME_1, NAME_2) |> dplyr::summarize(mean_yield = mean(grain_yield_kgpha)) |> dplyr::arrange(desc(mean_yield))
rmarkdown::paged_table(avg_yld_zone, options=NULL) 
```

Within and between zone and region Yield difference
```{r}
wheat_filtered |> 
ggplot(aes(y = grain_yield_kgpha, fill = NAME_2))+
  geom_boxplot(width = 0.2, outlier.colour = 'red')+
  facet_wrap(~NAME_1)+
  theme_bw()+
  theme(axis.text.x = element_blank(), legend.position = "none")+
   guides(fill = guide_legend(nrow = 4))
```

The average yield by district
```{r}
avg_yld_dist <- wheat_filtered |> dplyr::group_by(location) |> dplyr::summarize(mean_yield = mean(grain_yield_kgpha)) |> dplyr::arrange(desc(mean_yield))
rmarkdown::paged_table(avg_yld_dist, options=NULL) 
```

Few Observations by year
```{r}
obs_few_year <- wheat_filtered |> dplyr::group_by(year) |> dplyr::summarise(n = n()) |> dplyr::arrange(n)
rmarkdown::paged_table(obs_few_year, options=NULL) 
```

Observations by year
```{r}
few_year <- wheat_filtered |> dplyr::filter(year %in% c(1992, 1990, 1991, 2008, 2009, 2010)) |> 
  dplyr::group_by(year, trial_id) |> dplyr::summarise(n = n()) |> dplyr::arrange(n)
few_year
```

The data for years with observations are sourced from literature, isolated students work for PhD or MSc
```{r}
wheat_filtered |> dplyr::filter(year %in% c(1992, 1990, 1991, 2008, 2009, 2010)) |>
  ggplot(aes(x = treatment_id, y = grain_yield_kgpha, color = source))+
  geom_point(size = 2)+
  facet_grid(NAME_2 ~ year)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6))
```

## 21. Interactive map for yield  
```{r}
# Set tmap to view mode
eth_reg <- geodata::gadm(path = ".", country = "Eth", level = 1) |> st_as_sf()
wheat_filtered2 <- wheat_filtered |> 
  terra::vect(geom = c('X', 'Y'), crs = 'epsg:4326') |> st_as_sf()
tmap_mode("view")
tm_shape(eth)+
  tm_borders(col = "black") +
tm_shape(eth_reg) +
  tm_borders(col = "gray")+
tm_shape(wheat_filtered2)+
  tm_bubbles(col = "grain_yield_kgpha",
          title.col="Grain Yield (kg/ha)", 
          popup.vars=c("Zone"="NAME_2","DIstrict"="NAME_3",
                       "treatment"="treatment_id",
                       "Grain Yield"="grain_yield_kgpha", 
                       "Trial ID"="trial_id"),
          popup.format=list(grain_yield_kgpha=list(digits=2))) 
```

From the above map there is no spatial pattern of grain yield.



## 22. final selected data

```{r}
wheat_filtered <- wheat_filtered |> dplyr::filter(grain_yield_kgpha <= 15000) 
wheat_filtered |> 
  ggplot(aes(x = source, y = grain_yield_kgpha, fill = source))+
  geom_boxplot()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), legend.position = 'none')
```

<!-- Let's check the above dominant outliers between 7.5ton/ha and 15ton/ha to investigate whether these are outliers or not using location, year and source (trial id) -->

<!-- ```{r} -->
<!-- unq_tr_id <- wheat_filtered |> dplyr::filter(grain_yield_kgpha >= 7500 & grain_yield_kgpha <= 15000) -->
<!-- unq_tr_id <- unique(unq_tr_id$trial_id2) -->
<!-- unq_tr_id2 <- wheat_filtered |> dplyr::filter(grain_yield_kgpha <= 15000 & source %in% c("ICRISAT_ETH_Africa Rising", "ATA_Balanced_Fertilizer_Trial", "SG 2000") & year %in% c(2014, 2015, 2016) & trial_id %in% unq_tr_id) -->

<!--  unq_tr_id2 |>  -->
<!--   ggplot(aes(x = trial_id2, y = grain_yield_kgpha, fill = source))+ -->
<!--   geom_boxplot()+ -->
<!--   facet_wrap(~year, scales = "free")+ -->
<!--   theme_bw()+ -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8), legend.position = 'none') -->
<!-- ``` -->

<!-- Nutrient use for the above yield data -->
<!-- ```{r} -->
<!-- col_high_yld <- wheat_filtered |> dplyr::mutate(cy = ifelse(trial_id %in% unq_tr_id2$trial_id, 'green', 'red')) -->
<!-- col_high_yld |>  -->
<!--   ggplot(aes(x = n_rate2, y = grain_yield_kgpha, color = cy))+ -->
<!--   geom_point()+ -->
<!--   theme_bw()+ -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8), legend.position = 'none') -->
<!-- ``` -->

<!-- Nutrient use for the above yield data -->
<!-- ```{r} -->
<!-- col_high_yld <- wheat_filtered |> dplyr::mutate(cy = ifelse( trial_id %in% unq_tr_id2$trial_id, 'green', 'red')) -->
<!-- col_high_yld |>  -->
<!--   ggplot(aes(x = p_rate2, y = grain_yield_kgpha, color = cy))+ -->
<!--   geom_point()+ -->
<!--   theme_bw()+ -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8), legend.position = 'none') -->
<!-- ``` -->

<!-- For the two trials with lower yield above, we checked the treatment by yield.   -->
<!-- ```{r} -->
<!-- unq_tr_id2 |> dplyr::filter(grain_yield_kgpha <= 2500 & year %in% c(2014, 2015)) |>  -->
<!--   ggplot(aes(x = treatment_id, y = grain_yield_kgpha))+ -->
<!--   geom_point(color = "blue")+ -->
<!--   facet_wrap(~trial_id)+ -->
<!--   theme_bw() -->
<!-- ``` -->


## 23. Yield from the 0:0:0 treatment, the control by Zone and Region
It can help to understand regions with high baseline yield
```{r}
wheat_filtered |> dplyr::filter(treatment_id == "0_0_0") |>
ggplot(aes(y = grain_yield_kgpha, x = year, fill = as.factor(year)))+
  geom_boxplot()+
  ylab("yield")+
  facet_wrap(~NAME_1)+
  theme_bw()+
  theme(axis.text.x = element_blank(), axis.title.x =  element_blank())
```

```{r, fig.height=10, fig.width=13}
wheat_filtered |> dplyr::filter(treatment_id == "0_0_0" & year %in% c(2020,2021) & NAME_1 %in% c('Amhara', 'Oromia', 'SNNP')) |>
ggplot(aes(y = grain_yield_kgpha, x = NAME_2, fill = as.factor(year)))+
  geom_boxplot()+
  ylab("yield")+
  facet_wrap(~source, scale = 'free_x')+
  theme_bw()+
  xlab("Zone")+
  theme(axis.text.x = element_text(size = 14, angle = 15, hjust = 1), strip.text.x = element_text(size = 10))
```

First we select all trials with control, then we subset where control yield >=3.5ton/ha, we took all treatments of those trials and check the yield distribution. A box plot for those high control trials

<!-- The assumption is, the yield for other treatments must be bigger than the control yield.  -->
```{r, fig.height=10, fig.width=13}
ctrl_high_yld <- wheat_filtered |> dplyr::filter(treatment_id == "0_0_0"  & grain_yield_kgpha >= 3500 & grain_yield_kgpha <= 15000) 
wheat_filtered |> dplyr::filter(trial_id %in% unique(ctrl_high_yld$trial_id)) |>
ggplot(aes(y = grain_yield_kgpha, x = NAME_2, fill = as.factor(year)))+
  geom_boxplot()+
  geom_abline(slope = 0, intercept = 3500)+
  ylab("yield")+
  facet_wrap(~source, scale = 'free_x')+
  theme_bw()+
  xlab("Zone")+
  theme(axis.text.x = element_text(size = 14, angle = 15, hjust = 1), strip.text.x = element_text(size = 14))
```

## 24. Yield Effect

Reference treatment need to be selected within every trial as there are different NPK rates tested.
```{r, warning=FALSE}
options(dplyr.summarise.inform = FALSE)
unique_tr_id <- unique(wheat_filtered$trial_id)
yield_diff_mean <- data.frame()
for(i in 1:length(unique_tr_id)){
  unique_tr_yld <- wheat_filtered |> dplyr::filter(trial_id ==  unique_tr_id[i])
  max_n <- max(unique_tr_yld$n_rate2)
  max_p <- max(unique_tr_yld$p_rate2)
  max_k <- max(unique_tr_yld$k_rate2)
  unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p & k_rate2 == max_k)
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p)
  } 
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n)
  }
  if(nrow(unique_tr_yld2) > 1){
      ref_yld <- median(unique_tr_yld2$grain_yield_kgpha)
  }else{
      ref_yld <- unique_tr_yld2$grain_yield_kgpha
  }
unique_tr_yld <- unique_tr_yld |> dplyr::mutate(yield_diff = ref_yld - grain_yield_kgpha)
yield_diff_mean <- rbind(yield_diff_mean, unique_tr_yld)
}
saveRDS(yield_diff_mean, "~/shared-data/Data/Wheat/fieldData/yield_diff.rds")
```

Scatter plot of the yield difference: there are quite some negative yield effects in the data
```{r}
yield_diff_mean <- yield_diff_mean |> dplyr::mutate(ref_trt = ifelse(yield_diff == 0, T, F))
ref_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff == 0) |> dplyr::select(trial_id, yield_diff, treatment_id, grain_yield_kgpha)
other_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff != 0) |> dplyr::select(trial_id, yield_diff, treatment_id, grain_yield_kgpha)
all_diff_data <- ref_yld_data |> dplyr::inner_join(other_yld_data, by = "trial_id")

all_diff_data  |> ggplot(aes(y = grain_yield_kgpha.x, x = grain_yield_kgpha.y))+
  geom_point()+
  ylab("reference yield")+
  xlab("other treatements yield")+
  geom_abline(intercept = 0, slope = 1, color = 'blue')+
  xlim(0, 10000)+
  ylim(0, 10000)+
  theme_bw()
  #facet_wrap(~year)
```

## 25. Linear mixed effects model to reduce the ranomd noise in the data

```{r}
#create variables to deal with scale issues:
wheat_filtered$year <- as.factor(wheat_filtered$year)
wheat_filtered$source <- as.factor(wheat_filtered$source)
wheat_filtered$location <- as.factor(wheat_filtered$location)
wheat_filtered$trial_id <- as.factor(wheat_filtered$trial_id)
wheat_filtered$NAME_2 <- as.factor(wheat_filtered$NAME_2)

wheat_filtered$N100 <- wheat_filtered$n_rate2/100
wheat_filtered$P100 <- wheat_filtered$p_rate2/100
wheat_filtered$K100 <- wheat_filtered$k_rate2/100

fita <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + p_rate2 + k_rate2 + year + (1|trial_id), data = wheat_filtered)

fitb <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + p_rate2 + k_rate2 + year + (1|trial_id) + source, data = wheat_filtered)
anova(fita, fitb)

fitc <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + p_rate2 + k_rate2 + year + NAME_2 + (1|trial_id) + source, data = wheat_filtered)
anova(fitb, fitc)

fitd <- lmer(sqrt(grain_yield_kgpha) ~ n_rate2 + p_rate2 + k_rate2 + (1|year) + NAME_2 + (1|trial_id) + source, data = wheat_filtered)
anova(fitc, fitd)

fite <- update(fitc, . ~ . + N100:P100 + N100:K100 + P100:K100 + N100:P100:K100)
anova(fitc, fite)

fitf <- update(fite, . ~ . +(0 + N100|trial_id) +(0 + P100|trial_id) +(0 + K100|trial_id))
anova(fite, fitf)
#r.squaredGLMM(fitf) 

## fitting the model with RMLE
fitg <- update(fite, . ~ . +(0 + N100|trial_id) +(0 + P100|trial_id) +(0 + K100|trial_id))
wheat_filtered$blup <- predict(fitg, wheat_filtered)**2
```


## 26. compute the yield effect using the BLUP from the linear mixed effects model
```{r, warning=FALSE}
options(dplyr.summarise.inform = FALSE)
unique_tr_id <- unique(wheat_filtered$trial_id)
yield_diff_mean <- data.frame()
for(i in 1:length(unique_tr_id)){
  unique_tr_yld <- wheat_filtered |> dplyr::filter(trial_id ==  unique_tr_id[i])
  max_n <- max(unique_tr_yld$n_rate2)
  max_p <- max(unique_tr_yld$p_rate2)
  max_k <- max(unique_tr_yld$k_rate2)
  unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p & k_rate2 == max_k)
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n &  p_rate2 == max_p)
  } 
  if(nrow(unique_tr_yld2) == 0){
      unique_tr_yld2 <- unique_tr_yld |> dplyr::filter(n_rate2 == max_n)
  }
  if(nrow(unique_tr_yld2) > 1){
      ref_yld <- median(unique_tr_yld2$blup)
  }else{
      ref_yld <- unique_tr_yld2$blup
  }
unique_tr_yld <- unique_tr_yld |> dplyr::mutate(yield_diff = ref_yld - blup)
yield_diff_mean <- rbind(yield_diff_mean, unique_tr_yld)
}

```

Scatter plot of the yield difference with BLUPS
```{r}
yield_diff_mean <- yield_diff_mean |> dplyr::mutate(ref_trt = ifelse(yield_diff == 0, T, F))
ref_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff == 0) |> dplyr::select(trial_id, yield_diff, treatment_id, blup)
other_yld_data <- yield_diff_mean |> dplyr::filter(yield_diff != 0) |> dplyr::select(trial_id, yield_diff, treatment_id, blup)
all_diff_data <- ref_yld_data |> dplyr::inner_join(other_yld_data, by = "trial_id")

yield_diff_mean <- unique(yield_diff_mean[, c("trial_id","source", "year","NAME_1","NAME_2","NAME_3", "long2", "lat2","n_rate2", "p_rate2", "k_rate2", "treatment_id","ref_trt","grain_yield_kgpha","blup", "yield_diff")])

saveRDS(yield_diff_mean, "~/shared-data/Data/Wheat/fieldData/yield_diff_blup.rds")



all_diff_data  |> ggplot(aes(y = blup.x, x = blup.y))+
  geom_point()+
  ylab("reference yield")+
  xlab("other treatements yield")+
  geom_abline(intercept = 0, slope = 1, color = 'blue')+
  xlim(0, 10000)+
  ylim(0, 10000)+
  theme_bw()
  #facet_wrap(~year)
```


